---
title: "Characterising newly made gene sets"
author: "The Gene Signature Commons Group"
date: "`r Sys.Date()`"
output:
  html_document
---
Source: https://github.com/markziemann/gene_sig_commons

## Background

In this analysis we will be characterising the new gene sets.

## Libraries

```{r, libs}
suppressPackageStartupMessages({
  library("getDEE2")
  library("mitch")
  library("triwise")
  library("dplyr")
  library("gplots")
  library("reshape2")
  library("network")
})
```

## Reference gene list

This is the background - all genes in the "universe"

```{r,genelist}
# get universe of gene names and biotypes
if (! file.exists("hs.gtf.gz") ) {
  download.file("ftp://ftp.ensembl.org/pub/release-90/gtf/homo_sapiens/Homo_sapiens.GRCh38.90.gtf.gz",destfile= "hs.gtf.gz")
}

g<-read.table("hs.gtf.gz",sep="\t")
g <- g[grep("gene",g$V3),9] 
universe <- sapply(strsplit(g," "),"[[",6)
universe <- gsub(";","",universe)

biotypes <- sapply(strsplit(g," "),"[[",10)
biotypes <- gsub("_"," ",biotypes)
biotypes <- gsub(";","",biotypes)

biotypes_df <- data.frame(universe,biotypes)

universe <- unique(universe)

mytable <- table(biotypes)
mytable <- mytable[order(mytable)]
mytable <- mytable[which(mytable>100)]

par(mar=c(3,13,1,1)); barplot(mytable,horiz=TRUE,las=2,cex.names = 0.6,cex.axis = 0.6, 
  main="number of genes in each biotype class",xlim = c(0,20000)) ;grid()

# Take note how many transcripts are protein coding versus non-protein coding
paste("protein coding:",length(which(biotypes=="protein coding")))
paste("non-protein coding:",length(which(biotypes!="protein coding")))
```


## Curated gene sets

Here I download the current Reactome gene set library.

```{r,reactome}
download.file("https://reactome.org/download/current/ReactomePathways.gmt.zip", destfile="ReactomePathways.gmt.zip")
unzip("ReactomePathways.gmt.zip")
reactome <- gmt_import("ReactomePathways.gmt")
str(head(reactome))

reactome_genes <- unique(unname(unlist(reactome)))
head(reactome_genes)
head(universe)
length(intersect(reactome_genes,universe))
v1 <- list("Reactome"=reactome_genes, "Ensembl universe"=universe)
library("eulerr")
plot(euler(v1),quantities = TRUE)
length(reactome_genes)/length(universe)*100

prot <- biotypes_df[which(biotypes_df$biotypes=="protein coding"),1] 
prot <- unique(prot)
nprot <- biotypes_df[which(biotypes_df$biotypes!="protein coding"),1]
nprot <- unique(nprot)
intersect(prot,nprot)
nprot <-   setdiff(nprot,prot)
v1 <- list("Reactome"=reactome_genes, "protein coding"=prot, "non-protein coding"=nprot)
plot(euler(v1),quantities = TRUE)
```
Now let's do the same with GO sets downloaded from Ensembl biomart.

```{r,go}
go <- read.table("biomart_2020-09-26.txt.gz",sep="\t", fill = TRUE)
colnames(go) <- go[1,]
go <- go[2:nrow(go),]
go <- go[which(go$`GO term accession`!=""),]
go_genes <- unique(go$`Gene name`)
length(unique(go$`GO term accession`))
length(go_genes)
head(go)
v1 <- list("GO"=go_genes, "protein coding"=prot, "non-protein coding"=nprot)
plot(euler(v1),quantities = TRUE)
```
Human phenotype ontology
Downloaded from https://hpo.jax.org/app/download/annotation

```{r,hpo}
hpo <- read.table("phenotype_to_genes.txt",sep="\t")
head(hpo)
hpo_genes <- unique(hpo$V4)
v1 <- list("HPO"=hpo_genes, "protein coding"=prot, "non-protein coding"=nprot)
plot(euler(v1),quantities = TRUE)
```
MSigDB
```{r,msigdb}
msigdb <- gmt_import("msigdb.v7.2.symbols.gmt")
msigdb_genes <- unique(unname(unlist(msigdb)))
v1 <- list("MSigDB"=msigdb_genes,"protein coding"=prot, "non-protein coding"=nprot)
plot(euler(v1),quantities = TRUE)
```


## Defining the overrepresentation analysis function

Here I'm using the triwise::testEnrichment function which uses a Fisher test under the hood.

Reference
van de Laar L, Saelens W, De Prijck S, Martens L, Scott CL, Van Isterdael G, Hoffmann E, Beyaert R, Saeys Y, Lambrecht BN, Guilliams M. Yolk Sac Macrophages, Fetal Liver, and Adult Monocytes Can Colonize an Empty Niche and Develop into Functional Tissue-Resident Macrophages. Immunity. 2016 Apr 19;44(4):755-68. doi: 10.1016/j.immuni.2016.02.017. Epub 2016 Mar 15. PMID: 26992565.

```{r,orafunc}
ora <- function(gs,genesets,universe){ 
  res <- triwise::testEnrichment(gs, genesets, universe, minknown = 3, mindiffexp = 2, maxknown = 5000)
  if( nrow(res)>0 ) {
    res <- res[order(res$pval),]
    res <- head(res,5)
    hits <- log(res[which(res$qval<0.01),2])
    names(hits) <- res[which(res$qval<0.01),4]
    return(hits)
  } else {
    return(0)
  }
}
```

Here I'm declaring a function that converts list of vectors to a network diagram.

There are 2x edges than nodes.

Only the edges with highest similarity are retained, as per jaccard.

```{r,gsnet}
gs2net <- function(gset){
  mydf <- bind_rows(lapply(gset, as.data.frame.list))
  rownames(mydf) <- names(gset)
  mydf[1:5,1:6]
  #mydf[is.na(mydf)] <- 0
  j <- apply(mydf,1,function(x) {
    apply(mydf,1,function(y) {
      length(intersect(x,y) ) / length(union(x,y))
    })
  })
  j[lower.tri(j)] <- NA
  j[lower.tri(j,diag=TRUE)] <- 0
  jl <- melt(j)
  jl <- jl[which(jl$Var1 != jl$Var2),]
  jl <- jl[which(jl$value != 1),]
  jl <- jl[order(-jl$value),]
  jl <- head(jl,length(gset)*2)
  jl$edgeSize = with(jl, jl$value/sum(jl$value))
  jlnet = network(jl[,1:2])
  plot(jlnet, displaylabels = TRUE, label.col = "steelblue",
       edge.lwd = c(jl$edgeSize) * 100,
       arrowhead.cex = 0,
       label.cex = 1, vertex.border = "white",
       vertex.col = "blue", edge.col = rgb(0, 0, 0, alpha = 0.5))
}
```


## Epilepsy

Let's take a look at the epilepsy gene sets.

```{r,epilepsy,fig.width=10,fig.height=8}
epi <- gmt_import("epilepsy.gmt")
epi <- epi[which(!duplicated(names(epi)))]
setnames <- names(epi)
names(epi) <-paste("GS",1:length(epi))
numgenes <- unlist(lapply(epi,length))
hist(numgenes,breaks = 15, xlab="number of genes per set",main="Epilepsy gene sets")

gs2net(epi)

# classes
mytable <- lapply(epi,function(x) {
  table(biotypes_df[which(x %in% biotypes_df$universe),2])
})
mydf <- bind_rows(lapply(mytable, as.data.frame.list))
#mydf[is.na(mydf)] <- 0
rownames(mydf) <- names(epi)
mydf_n <- mydf/rowSums(mydf)
mydf <- mydf[,head(order(-colSums(mydf_n)),8)]
colfunc <- colorRampPalette(c("white", "yellow","orange","red","darkred"))(n=25)
heatmap.2(as.matrix(mydf),col=colfunc,trace="none",margin=c(10,15),cexCol = 0.9, cexRow = 0.7,scale="row",dendrogram = "none",main="Epilepsy gene sets by biotype",na.color = "white")

mydf_n[is.na(mydf_n)] <- 0
mytable <-colSums(mydf_n)
mytable <- mytable/sum(mytable)*100
mytable<-mytable[order(mytable)]
par(mar=c(3,13,1,1)); barplot(mytable,horiz=TRUE,las=2,cex.names = 0.7,cex.axis = 0.8, 
  main="Overall biotype class representation (%)" ) ;grid()

# ORA
res <- lapply(epi, function(x) { ora(gs=x,genesets=reactome, universe=universe)  }  )
res <- res[which(lapply(res,length)>0)]
myres <- bind_rows(lapply(res, as.data.frame.list))
myres[is.na(myres)] <- 0
rownames(myres) <- names(res)
colfunc <- colorRampPalette(c("white", "lightblue","blue", "darkblue"))(n=25)
heatmap.2(as.matrix(myres),col=colfunc,trace="none",margin=c(10,15),cexRow = 0.7,dendrogram = "none",main="Epilepsy gene sets by Reactome enrichment")
```


## Diabetes

Let's take a look at the diabetes gene sets.

```{r,diabetes,fig.width=8,fig.height=8}
diab <- gmt_import("diabetes.gmt")
diab <- diab[which(!duplicated(names(diab)))]
setnames <- names(diab)
names(diab) <-paste("GS",1:length(diab))
numgenes <- unlist(lapply(diab,length))
hist(numgenes,breaks = 15, xlab="number of genes per set",main="Diabetes gene sets")

gs2net(diab)

# classes
mytable <- lapply(diab,function(x) {
  table(biotypes_df[which(x %in% biotypes_df$universe),2])
})
mydf <- bind_rows(lapply(mytable, as.data.frame.list))
rownames(mydf) <- names(diab)
mydf_n <- mydf/rowSums(mydf)
mydf <- mydf[,head(order(-colSums(mydf_n)),8)]
colfunc <- colorRampPalette(c("white", "yellow","orange","red","darkred"))(n=25)
heatmap.2(as.matrix(mydf),col=colfunc,trace="none",margin=c(10,30),cexCol = 0.6, cexRow = 0.2 ,scale="row",dendrogram = "none",main="Diabetes gene sets by biotype",na.color = "white")

mydf_n[is.na(mydf_n)] <- 0
mytable <-colSums(mydf_n)
mytable <- mytable/sum(mytable)*100
mytable<-mytable[order(mytable)]
par(mar=c(3,13,1,1)); barplot(mytable,horiz=TRUE,las=2,cex.names = 0.7,cex.axis = 0.8, 
  main="Overall biotype class representation (%)" ) ;grid()

# reactome enrichment
res <- lapply(diab, function(x) { ora(gs=x,genesets=reactome, universe=universe)  }  )
res <- res[which(lapply(res,length)>0)]
myres <- bind_rows(lapply(res, as.data.frame.list))
myres[is.na(myres)] <- 0
rownames(myres) <- names(res)
colfunc <- colorRampPalette(c("white", "lightblue","blue", "darkblue"))(n=25)
heatmap.2(as.matrix(myres),col=colfunc,trace="none",margin=c(10,15),cexRow = 0.2, cexCol = 0.3 ,dendrogram = "none",main="Diabetes gene sets by Reactome enrichment")
```


## Heart disease

Let's take a look at the heart disease gene sets.

```{r,heartdisease,fig.width=8,fig.height=8}
hd <- gmt_import("heartdisease.gmt")
hd <- hd[which(!duplicated(names(hd)))]
setnames <- names(hd)
names(hd) <-paste("GS",1:length(hd))
numgenes <- unlist(lapply(hd,length))
hist(numgenes,breaks = 15, xlab="number of genes per set",main="Heart disease gene sets")

gs2net(hd)

# classes
mytable <- lapply(hd,function(x) {
  table(biotypes_df[which(x %in% biotypes_df$universe),2])
})
mydf <- bind_rows(lapply(mytable, as.data.frame.list))
rownames(mydf) <- names(hd)
mydf_n <- mydf/rowSums(mydf)
mydf <- mydf[,head(order(-colSums(mydf_n)),8)]
colfunc <- colorRampPalette(c("white", "yellow","orange","red","darkred"))(n=25)
heatmap.2(as.matrix(mydf),col=colfunc,trace="none",margin=c(10,30),cexCol = 0.8, cexRow = 0.5,scale="row",dendrogram = "none",main="Heart disease gene sets by biotype",na.color = "white")

mydf_n[is.na(mydf_n)] <- 0
mytable <-colSums(mydf_n)
mytable <- mytable/sum(mytable)*100
mytable<-mytable[order(mytable)]
par(mar=c(3,13,1,1)); barplot(mytable,horiz=TRUE,las=2,cex.names = 0.7,cex.axis = 0.8, 
  main="Overall biotype class representation (%)" ) ;grid()

# reactome enrichment
res <- lapply(hd, function(x) { ora(gs=x,genesets=reactome, universe=universe)  }  )
res <- res[which(lapply(res,length)>0)]
myres <- bind_rows(lapply(res, as.data.frame.list))
myres[is.na(myres)] <- 0
rownames(myres) <- names(res)
colfunc <- colorRampPalette(c("white", "lightblue","blue", "darkblue"))(n=25)
heatmap.2(as.matrix(myres),col=colfunc,trace="none",margin=c(10,15),cexRow = 0.5, cexCol = 0.5, dendrogram = "none",main="Heart disease gene sets by Reactome enrichment")
```

## SARS, MERS and SARS-CoV-2

Let's take a look at these virus infection gene sets.

```{r,sars,fig.width=8,fig.height=10}
sars <- gmt_import("SARS_MERS.gmt")
sars <- sars[which(!duplicated(names(sars)))]
setnames <- names(sars)
names(sars) <-paste("GS",1:length(sars))
numgenes <- unlist(lapply(sars,length))
hist(numgenes,breaks = 15, xlab="number of genes per set",main="SARS, MERS and SARS-CoV-2 gene sets")

gs2net(sars)

# classes
mytable <- lapply(sars,function(x) {
  table(biotypes_df[which(x %in% biotypes_df$universe),2])
})
mydf <- bind_rows(lapply(mytable, as.data.frame.list))
rownames(mydf) <- names(sars)
mydf_n <- mydf/rowSums(mydf)
mydf <- mydf[,head(order(-colSums(mydf_n)),8)]
colfunc <- colorRampPalette(c("white", "yellow","orange","red","darkred"))(n=25)
heatmap.2(as.matrix(mydf),col=colfunc,trace="none",margin=c(10,30),cexCol = 0.8, cexRow = 0.5,scale="row",dendrogram = "none",main="SARS, MERS and SARS-CoV2 gene sets by biotype",na.color = "white")

mydf_n[is.na(mydf_n)] <- 0
mytable <-colSums(mydf_n)
mytable <- mytable/sum(mytable)*100
mytable<-mytable[order(mytable)]
par(mar=c(3,13,1,1)); barplot(mytable,horiz=TRUE,las=2,cex.names = 0.7,cex.axis = 0.8, 
  main="Overall biotype class representation (%)" ) ;grid()

# Reactome enrichment
res <- lapply(sars, function(x) { ora(gs=x,genesets=reactome, universe=universe)  }  )
res <- res[which(lapply(res,length)>0)]
myres <- bind_rows(lapply(res, as.data.frame.list))
myres[is.na(myres)] <- 0
rownames(myres) <- names(res)
colfunc <- colorRampPalette(c("white", "lightblue","blue", "darkblue"))(n=25)
heatmap.2(as.matrix(myres),col=colfunc,trace="none",margin=c(10,15),cexRow = 0.5, cexCol = 0.4,dendrogram = "none",main="SARS, MERS and SARS-CoV-2 gene sets genes by Reactome enrichment")
```


## Session information

```{r sessioninfo}
sessionInfo()
```

